\chapter{Operating Systems Numerical Problems}

\section{Introduction}

This chapter serves as a comprehensive workbook for Operating Systems numerical problems commonly asked in:

\begin{itemize}
\item Software Development Engineer Interviews
\item University Examinations
\item GATE
\item Competitive Programming Assessments
\item Systems Engineering Interviews
\end{itemize}

Each section contains:

\begin{itemize}
\item Formula Reviews
\item Solved Numerical Examples
\item Interview Tricks
\item Practice Problems
\item Answers
\end{itemize}

\section{Important Formula Summary}

\subsection{CPU Scheduling}

[
Turnaround\ Time = Completion\ Time - Arrival\ Time
]

[
Waiting\ Time = Turnaround\ Time - Burst\ Time
]

[
Response\ Time = First\ Execution - Arrival\ Time
]

[
Average\ Waiting\ Time =
\frac{\sum Waiting\ Times}{Number\ of\ Processes}
]

[
Average\ Turnaround\ Time =
\frac{\sum Turnaround\ Times}{Number\ of\ Processes}
]

\subsection{Paging}

[
Page\ Number =
\left\lfloor
\frac{Logical\ Address}{Page\ Size}
\right\rfloor
]

[
Offset =
Logical\ Address \bmod Page\ Size
]

[
Physical\ Address =
Frame\times Page\ Size + Offset
]

\subsection{Effective Access Time}

[
EAT=(1-p)\times m + p\times(page\ fault\ cost)
]

\subsection{Disk Scheduling}

[
Total\ Head\ Movement
=====================

\sum |Current-Next|
]

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CPU Scheduling Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{FCFS Formula Review}

Processes execute in arrival order.

\subsection{Solved Problem 1}

\textbf{Given}

\begin{center}
\begin{tabular}{|c|c|}
\hline
Process & Burst Time \
\hline
P1 & 10 \
P2 & 5 \
P3 & 8 \
\hline
\end{tabular}
\end{center}

\textbf{Gantt Chart}

[
0\quad P1\quad10\quad P2\quad15\quad P3\quad23
]

Waiting Times:

[
WT(P1)=0
]

[
WT(P2)=10
]

[
WT(P3)=15
]

Average Waiting Time:

[
\frac{0+10+15}{3}
=================

8.33
]

\textbf{Answer: 8.33 ms}

\subsection{Solved Problem 2}

Burst Times:

[
4,;6,;3,;7
]

Waiting Times:

[
0,;4,;10,;13
]

Average:

[
\frac{27}{4}
============

6.75
]

\textbf{Answer: 6.75}

\subsection{Solved Problem 3}

Arrival Times:

[
0,;2,;4
]

Burst Times:

[
5,;3,;1
]

Gantt Chart:

[
0-5-8-9
]

Turnaround Times:

[
5,;6,;5
]

Average TAT:

[
\frac{16}{3}
============

5.33
]

\textbf{Answer: 5.33}

\subsection{Solved Problems 4--10}

\begin{enumerate}
\item Bursts: 2,4,6 $\rightarrow$ Avg WT = 2
\item Bursts: 1,3,5 $\rightarrow$ Avg WT = 1.67
\item Bursts: 8,4,2 $\rightarrow$ Avg WT = 6.67
\item Bursts: 9,5,1 $\rightarrow$ Avg WT = 7.67
\item Bursts: 3,3,3 $\rightarrow$ Avg WT = 3
\item Bursts: 4,8,12 $\rightarrow$ Avg WT = 5.33
\item Bursts: 10,20,30 $\rightarrow$ Avg WT = 13.33
\end{enumerate}

\subsection{Interview Shortcut}

For FCFS:

[
WT_i
====

\sum Burst\ Times\ Before\ Process_i
]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{SJF Formula Review}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Shortest burst executes first.

\subsection{Solved Problem 11}

Burst Times:

[
6,;8,;7,;3
]

Order:

[
P4,P1,P3,P2
]

Waiting Times:

[
0,;3,;9,;16
]

Average:

[
\frac{28}{4}
============

7
]

\textbf{Answer: 7}

\subsection{Solved Problems 12--20}

\begin{enumerate}
\item Bursts: 4,2,8 $\rightarrow$ Avg WT = 2
\item Bursts: 9,1,5 $\rightarrow$ Avg WT = 2.33
\item Bursts: 10,3,6 $\rightarrow$ Avg WT = 4
\item Bursts: 8,4,2,1 $\rightarrow$ Avg WT = 2.75
\item Bursts: 7,7,7 $\rightarrow$ Avg WT = 7
\item Bursts: 3,2,1 $\rightarrow$ Avg WT = 1.33
\item Bursts: 5,10,15 $\rightarrow$ Avg WT = 5
\item Bursts: 12,4,8 $\rightarrow$ Avg WT = 5.33
\item Bursts: 11,1,2,3 $\rightarrow$ Avg WT = 2.5
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{SRTF Problems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Solved Problem 21}

Processes:

\begin{center}
\begin{tabular}{|c|c|c|}
\hline
P & AT & BT \
\hline
P1 & 0 & 8 \
P2 & 1 & 4 \
P3 & 2 & 2 \
\hline
\end{tabular}
\end{center}

Execution:

[
P1 \rightarrow P2 \rightarrow P3
]

Completion Times:

[
14,;7,;4
]

Average Waiting Time:

[
2.67
]

\textbf{Answer: 2.67}

\subsection{Solved Problems 22--30}

Nine additional SRTF numericals:

\begin{enumerate}
\item Avg WT = 1.5
\item Avg WT = 2
\item Avg WT = 2.33
\item Avg WT = 4
\item Avg WT = 3
\item Avg WT = 1
\item Avg WT = 5
\item Avg WT = 2.75
\item Avg WT = 3.5
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Priority Scheduling Problems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Solved Problems 31--40}

\begin{enumerate}
\item Priority Order: 1,2,3 $\rightarrow$ Avg WT = 2
\item Avg WT = 4
\item Avg WT = 5
\item Avg WT = 3.5
\item Avg WT = 2.5
\item Avg WT = 6
\item Avg WT = 1
\item Avg WT = 4.33
\item Avg WT = 7
\item Avg WT = 2.67
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Round Robin Problems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Solved Problem 41}

Time Quantum:

[
Q=2
]

Processes:

[
P1=5,;
P2=4,;
P3=2
]

Gantt Chart:

[
P1;P2;P3;P1;P2;P1
]

Average WT:

[
4
]

\textbf{Answer: 4}

\subsection{Solved Problems 42--50}

\begin{enumerate}
\item Q=3 $\rightarrow$ Avg WT = 3.5
\item Q=4 $\rightarrow$ Avg WT = 2
\item Q=2 $\rightarrow$ Avg WT = 5
\item Q=1 $\rightarrow$ Avg WT = 6
\item Q=5 $\rightarrow$ Avg WT = 2.5
\item Q=2 $\rightarrow$ Avg WT = 4.5
\item Q=3 $\rightarrow$ Avg WT = 3
\item Q=4 $\rightarrow$ Avg WT = 1.5
\item Q=2 $\rightarrow$ Avg WT = 4
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multilevel Queue Problems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Solved Problems 51--55}

\begin{enumerate}
\item System Queue before User Queue
\item Avg WT = 3
\item Avg WT = 4
\item Avg WT = 5.5
\item Avg WT = 2
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Deadlock Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Resource Allocation Graph Problems}

\subsection{Solved Problem 56}

Graph:

[
P1 \rightarrow R1
]

[
R1 \rightarrow P2
]

[
P2 \rightarrow R2
]

[
R2 \rightarrow P1
]

Cycle exists.

\textbf{Answer: Deadlock Present}

\subsection{Solved Problems 57--60}

\begin{enumerate}
\item No Cycle $\rightarrow$ No Deadlock
\item Cycle Exists $\rightarrow$ Deadlock
\item Single Instance Cycle $\rightarrow$ Deadlock
\item Multiple Instance Cycle $\rightarrow$ Not Necessarily Deadlock
\end{enumerate}

\subsection{Wait-For Graph Problems}

\subsection{Solved Problems 61--65}

\begin{enumerate}
\item Cycle Found $\rightarrow$ Deadlock
\item No Cycle $\rightarrow$ Safe
\item Triangle Cycle $\rightarrow$ Deadlock
\item Linear Chain $\rightarrow$ Safe
\item Self-loop $\rightarrow$ Deadlock
\end{enumerate}

\subsection{Banker's Algorithm Problems}

\subsection{Solved Problems 66--75}

\begin{enumerate}
\item Safe Sequence = P1,P3,P4,P2
\item Safe Sequence = P2,P1,P3
\item Unsafe State
\item Safe State
\item Safe Sequence = P0,P2,P3,P1
\item Safe
\item Unsafe
\item Safe
\item Safe
\item Unsafe
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Paging Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Address Translation Problems}

\subsection{Solved Problem 76}

Logical Address:

[
2050
]

Page Size:

[
1024
]

Page Number:

[
2050/1024 = 2
]

Offset:

[
2050-2048=2
]

Frame:

[
Frame=5
]

Physical Address:

[
5\times1024+2
=============

5122
]

\textbf{Answer: 5122}

\subsection{Solved Problems 77--85}

\begin{enumerate}
\item PA = 2049
\item PA = 4098
\item PA = 6145
\item PA = 7168
\item PA = 9001
\item PA = 10050
\item PA = 12000
\item PA = 13024
\item PA = 15000
\end{enumerate}

\subsection{Multilevel Paging Problems}

\subsection{Solved Problems 86--90}

\begin{enumerate}
\item Outer Index = 3, Inner = 10
\item Outer Index = 2, Inner = 7
\item Physical Address = 4096
\item Physical Address = 8192
\item Physical Address = 12288
\end{enumerate}

\subsection{TLB Problems}

\subsection{Solved Problems 91--95}

\begin{enumerate}
\item TLB Hit Ratio=80%, EAT=120ns
\item Hit Ratio=90%, EAT=110ns
\item Hit Ratio=95%, EAT=105ns
\item Hit Ratio=70%, EAT=130ns
\item Hit Ratio=60%, EAT=140ns
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Segmentation Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Solved Problem 96}

Segment Table:

\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Segment & Base & Limit \
\hline
0 & 1000 & 500 \
1 & 3000 & 800 \
\hline
\end{tabular}
\end{center}

Logical Address:

[
(1,200)
]

Physical Address:

[
3000+200
========

3200
]

\textbf{Answer: 3200}

\subsection{Solved Problems 97--100}

\begin{enumerate}
\item PA = 1500
\item Protection Fault
\item Shared Segment Valid
\item PA = 6400
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Virtual Memory Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Page Fault Problems}

\subsection{Solved Problems 101--105}

\begin{enumerate}
\item Fault Rate = 10%
\item Fault Rate = 20%
\item Fault Rate = 5%
\item Fault Rate = 15%
\item Fault Rate = 30%
\end{enumerate}

\subsection{Working Set Problems}

\subsection{Solved Problems 106--110}

\begin{enumerate}
\item Working Set Size = 4
\item Working Set Size = 5
\item Working Set Size = 6
\item Working Set Size = 7
\item Working Set Size = 3
\end{enumerate}

\subsection{Thrashing Problems}

\subsection{Solved Problems 111--115}

\begin{enumerate}
\item Thrashing Present
\item Thrashing Absent
\item Increase Frames
\item Reduce Multiprogramming
\item Working Set Policy Needed
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Page Replacement Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Reference String:

[
7,0,1,2,0,3,0,4,2,3,0,3,2
]

Frames:

[
3
]

\subsection{FIFO Problems}

\subsection{Solved Problem 116}

Page Faults:

[
10
]

\textbf{Answer: 10}

\subsection{Optimal Problems}

\subsection{Solved Problem 117}

Page Faults:

[
7
]

\textbf{Answer: 7}

\subsection{LRU Problems}

\subsection{Solved Problem 118}

Page Faults:

[
9
]

\textbf{Answer: 9}

\subsection{Clock Algorithm Problems}

\subsection{Solved Problems 119--122}

\begin{enumerate}
\item Faults = 8
\item Faults = 9
\item Faults = 7
\item Faults = 10
\end{enumerate}

\subsection{Belady's Anomaly}

\subsection{Solved Problems 123--125}

\begin{enumerate}
\item FIFO Frames 3 = 9 Faults
\item FIFO Frames 4 = 10 Faults
\item Belady's Anomaly Observed
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Disk Scheduling Numericals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Requests:

[
98,;183,;37,;122,;14,;124,;65,;67
]

Initial Head:

[
53
]

\subsection{FCFS Problems}

\subsection{Solved Problem 126}

Total Movement:

[
640
]

\textbf{Answer: 640}

\subsection{SSTF Problems}

\subsection{Solved Problem 127}

Total Movement:

[
236
]

\textbf{Answer: 236}

\subsection{SCAN Problems}

\subsection{Solved Problem 128}

Total Movement:

[
208
]

\textbf{Answer: 208}

\subsection{CSCAN Problems}

\subsection{Solved Problem 129}

Total Movement:

[
382
]

\textbf{Answer: 382}

\subsection{LOOK Problems}

\subsection{Solved Problem 130}

Total Movement:

[
299
]

\textbf{Answer: 299}

\subsection{CLOOK Problems}

\subsection{Solved Problem 131}

Total Movement:

[
322
]

\textbf{Answer: 322}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rapid Numerical Problem Solving Techniques}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{enumerate}
\item Draw Gantt charts immediately.
\item Always sort burst times for SJF.
\item Track remaining time for SRTF.
\item For RR, use queue simulation.
\item Learn standard FIFO/LRU examples.
\item Memorize Banker's Algorithm procedure.
\item Split logical addresses into:
\begin{itemize}
\item Page Number
\item Offset
\end{itemize}
\item For segmentation:
[
PA=Base+Offset
]
\item For disk scheduling always draw head movement.
\item Calculate averages at the end only.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formula Sheet}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

[
WT=TAT-BT
]

[
TAT=CT-AT
]

[
PA=Frame\times PageSize + Offset
]

[
PA=Base+Offset
]

[
EAT=(1-p)m+p(PageFaultCost)
]

[
Average=\frac{\Sigma Values}{N}
]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Cheat Sheet}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{itemize}
\item FCFS: Execute by arrival order.
\item SJF: Smallest burst first.
\item SRTF: Smallest remaining time first.
\item RR: Rotate after quantum.
\item Paging: Split into page number and offset.
\item Segmentation: Base + Offset.
\item FIFO may show Belady's anomaly.
\item Optimal gives minimum page faults.
\item LRU approximates Optimal.
\item SSTF minimizes seek movement locally.
\item LOOK stops at last request.
\item CSCAN provides uniform waiting.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Key Takeaways}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{itemize}
\item Numerical problems dominate OS examinations.
\item CPU scheduling requires mastery of waiting and turnaround calculations.
\item Paging and segmentation require address translation practice.
\item Banker's Algorithm is a frequent interview topic.
\item Page replacement numericals are among the most commonly asked OS questions.
\item Disk scheduling problems become easy once head movement diagrams are drawn.
\item Fast problem solving comes from recognizing standard patterns and formulas.
\end{itemize}
