\chapter{Operating Systems Interview Questions}

\section{Introduction}

This chapter is a consolidated interview-preparation resource covering the most important Operating Systems topics frequently asked in software engineering interviews.

\section{Operating System Fundamentals}

\begin{enumerate}
\item \textbf{Interview Question:} What is an operating system?\
\textbf{Short Answer:} Resource manager and abstraction layer.\
\textbf{Detailed Explanation:} Manages CPU, memory, storage, and devices.\
\textbf{Common Follow-up Questions:} Kernel? System calls?\
\textbf{Interview Tips:} Mention abstraction and resource allocation.

\item \textbf{Interview Question:} What are system calls?\
\textbf{Short Answer:} Interface to kernel services.\
\textbf{Detailed Explanation:} User programs access OS functionality through them.\
\textbf{Common Follow-up Questions:} fork()? exec()?\
\textbf{Interview Tips:} Explain mode switch.

\item \textbf{Interview Question:} What is a kernel?
\item \textbf{Interview Question:} Monolithic vs microkernel?
\item \textbf{Interview Question:} User mode vs kernel mode?
\item \textbf{Interview Question:} What is a trap?
\item \textbf{Interview Question:} What is an interrupt?
\item \textbf{Interview Question:} What is DMA?
\item \textbf{Interview Question:} What is bootstrapping?
\item \textbf{Interview Question:} What is a context switch?
\end{enumerate}

\section{Processes}

\begin{enumerate}
\setcounter{enumi}{10}
\item Process lifecycle?
\item Process Control Block?
\item fork() vs exec()?
\item Zombie process?
\item Orphan process?
\item Process states?
\item Process creation?
\item Process termination?
\item Daemon process?
\item Copy-on-write during fork?
\end{enumerate}

\section{Threads}

\begin{enumerate}
\setcounter{enumi}{20}
\item What is a thread?
\item Process vs thread?
\item User vs kernel threads?
\item Thread states?
\item Thread pool?
\item Multithreading benefits?
\item Thread safety?
\item Reentrant functions?
\item Thread-local storage?
\item Why are threads lightweight?
\end{enumerate}

\section{CPU Scheduling}

\begin{enumerate}
\setcounter{enumi}{30}
\item Scheduling objective?
\item FCFS scheduling?
\item SJF scheduling?
\item SRTF scheduling?
\item Round Robin?
\item Priority scheduling?
\item Starvation?
\item Aging?
\item Throughput vs turnaround time?
\item Context-switch overhead?
\end{enumerate}

\section{Synchronization}

\begin{enumerate}
\setcounter{enumi}{40}
\item Race condition?
\item Critical section?
\item Mutual exclusion?
\item Progress condition?
\item Bounded waiting?
\item Peterson's algorithm?
\item Spinlock?
\item Semaphore?
\item Mutex?
\item Monitor?
\end{enumerate}

\section{Deadlocks}

\begin{enumerate}
\setcounter{enumi}{50}
\item What is deadlock?
\item Four deadlock conditions?
\item Mutual exclusion condition?
\item Hold and wait?
\item No preemption?
\item Circular wait?
\item Deadlock prevention?
\item Deadlock avoidance?
\item Deadlock detection?
\item Deadlock recovery?
\end{enumerate}

\section{Memory Management}

\begin{enumerate}
\setcounter{enumi}{60}
\item Logical vs physical memory?
\item Memory allocation?
\item Fragmentation?
\item Internal fragmentation?
\item External fragmentation?
\item Compaction?
\item Buddy allocator?
\item Slab allocator?
\item Heap management?
\item Free lists?
\end{enumerate}

\section{Virtual Memory}

\begin{enumerate}
\setcounter{enumi}{70}
\item Why virtual memory?
\item Virtual address space?
\item Demand paging?
\item Page fault?
\item Valid vs invalid reference?
\item Working set model?
\item Thrashing?
\item Swap space?
\item Copy-on-write?
\item Memory-mapped files?
\end{enumerate}

\section{Paging and Segmentation}

\begin{enumerate}
\setcounter{enumi}{80}
\item What is paging?
\item Page vs frame?
\item Page table?
\item TLB?
\item Effective access time?
\item Multilevel paging?
\item Inverted page table?
\item Hashed page table?
\item Segmentation?
\item Paging vs segmentation?
\end{enumerate}

\section{Page Replacement}

\begin{enumerate}
\setcounter{enumi}{90}
\item FIFO replacement?
\item Optimal replacement?
\item LRU replacement?
\item LFU replacement?
\item Second chance?
\item Clock algorithm?
\item Belady's anomaly?
\item Reference string?
\item Working set replacement?
\item Global vs local replacement?
\end{enumerate}

\section{File Systems}

\begin{enumerate}
\setcounter{enumi}{100}
\item File system purpose?
\item Inode?
\item Superblock?
\item File Control Block?
\item Contiguous allocation?
\item Linked allocation?
\item Indexed allocation?
\item Journaling?
\item ext4 architecture?
\item NTFS overview?
\end{enumerate}

\section{Disk Scheduling}

\begin{enumerate}
\setcounter{enumi}{110}
\item Disk scheduling objective?
\item FCFS disk scheduling?
\item SSTF?
\item SCAN?
\item C-SCAN?
\item LOOK?
\item C-LOOK?
\item Seek time?
\item Rotational latency?
\item Throughput optimization?
\end{enumerate}

\section{Linux Internals}

\begin{enumerate}
\setcounter{enumi}{120}
\item Linux architecture?
\item System call path?
\item proc filesystem?
\item sysfs?
\item CFS scheduler?
\item Signals?
\item Kernel modules?
\item systemd?
\item Linux boot process?
\item Kernel panic?
\end{enumerate}

\section{Containers and Virtualization}

\begin{enumerate}
\setcounter{enumi}{130}
\item Why virtualization?
\item Hypervisor?
\item Type 1 hypervisor?
\item Type 2 hypervisor?
\item Full virtualization?
\item Para-virtualization?
\item Containers?
\item Namespaces?
\item cgroups?
\item Docker vs VM?
\end{enumerate}

\section{Operating System Security}

\begin{enumerate}
\setcounter{enumi}{140}
\item Principle of least privilege?
\item Access control?
\item Authentication vs authorization?
\item Capability-based security?
\item Sandboxing?
\item ASLR?
\item Stack canaries?
\item Secure boot?
\item SELinux?
\item Privileged instructions?
\end{enumerate}

\section{High-Concurrency Systems}

\begin{enumerate}
\setcounter{enumi}{150}
\item What is high concurrency?
\item Thread-per-request model?
\item Event-driven architecture?
\item Reactor pattern?
\item Proactor pattern?
\item Non-blocking I/O?
\item Asynchronous I/O?
\item epoll?
\item Backpressure?
\item How does Nginx scale?
\end{enumerate}

\section{Top 25 Most Frequently Asked Questions}

\begin{enumerate}
\item Process vs Thread
\item Mutex vs Semaphore
\item Deadlock Conditions
\item What Happens During a Page Fault
\item Paging vs Segmentation
\item Virtual Memory
\item TLB
\item Context Switch
\item fork vs exec
\item Zombie Process
\item Orphan Process
\item CPU Scheduling Algorithms
\item Critical Section Problem
\item Peterson's Algorithm
\item Producer Consumer Problem
\item Readers Writers Problem
\item Dining Philosophers
\item Banker's Algorithm
\item LRU Page Replacement
\item Belady's Anomaly
\item Inode Structure
\item Linux Boot Process
\item System Calls
\item Docker vs VM
\item epoll vs select
\end{enumerate}

\section{Top 25 One-Line Rapid Revision Questions}

\begin{enumerate}
\item What is a process?
\item What is a thread?
\item What is a PCB?
\item What is a page?
\item What is a frame?
\item What is TLB?
\item What is DMA?
\item What is a trap?
\item What is an interrupt?
\item What is deadlock?
\item What is starvation?
\item What is aging?
\item What is a semaphore?
\item What is a mutex?
\item What is thrashing?
\item What is swapping?
\item What is journaling?
\item What is an inode?
\item What is CFS?
\item What is a namespace?
\item What is a cgroup?
\item What is epoll?
\item What is a reactor?
\item What is a hypervisor?
\item What is copy-on-write?
\end{enumerate}

\section{Top 25 Deep-Dive Questions}

\begin{enumerate}
\item Explain page-fault handling in detail.
\item Explain Linux process creation.
\item Explain Linux memory management.
\item Explain TLB shootdown.
\item Explain NUMA architecture.
\item Explain lock-free programming.
\item Explain compare-and-swap.
\item Explain monitor implementation.
\item Explain distributed locks.
\item Explain CFS scheduler internals.
\item Explain ext4 journaling.
\item Explain virtual memory in Linux.
\item Explain container isolation.
\item Explain namespaces and cgroups.
\item Explain epoll internals.
\item Explain event-driven architectures.
\item Explain context-switch costs.
\item Explain Banker's Algorithm.
\item Explain demand paging.
\item Explain working-set model.
\item Explain slab allocation.
\item Explain page replacement trade-offs.
\item Explain kernel-user transitions.
\item Explain memory ordering.
\item Explain high-concurrency server design.
\end{enumerate}

\section{Interview Preparation Strategy}

\begin{enumerate}
\item Master operating-system fundamentals.
\item Understand processes and threads thoroughly.
\item Practice synchronization problems.
\item Solve deadlock scenarios.
\item Learn memory management numericals.
\item Practice page replacement questions.
\item Understand Linux internals.
\item Learn virtualization and containers.
\item Connect OS concepts to system design.
\item Conduct mock interviews regularly.
\end{enumerate}

\section{Last-Minute Revision Advice}

\begin{itemize}
\item Revise Process vs Thread.
\item Revise Mutex vs Semaphore.
\item Memorize deadlock conditions.
\item Practice page-fault handling.
\item Revise TLB and paging.
\item Understand copy-on-write.
\item Review CPU scheduling algorithms.
\item Revise producer-consumer and readers-writers.
\item Review Linux process lifecycle.
\item Revise Docker and container isolation.
\item Review epoll and event-driven systems.
\item Practice concise interview explanations.
\end{itemize}

\section{Key Takeaways}

\begin{keytakeaways}
Operating Systems remain one of the highest-signal interview domains for software engineers. Focus on processes, threads, synchronization, deadlocks, memory management, virtual memory, Linux internals, containers, and high-concurrency architectures. Strong candidates connect operating-system fundamentals with modern backend systems, distributed systems, cloud infrastructure, and large-scale system design. Consistent practice of conceptual, numerical, and scenario-based questions is the most effective preparation strategy.
\end{keytakeaways}
